A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-14 latex/main.tex

\subsection*{Comparison with RIPPLES}

RIPPLES~\cite{Turakhia2022} is a method for identifying
recombination \emph{events} using a phylogeny inferred by UShER.
It is the only \emph{genealogical} method built for this purpose and
therefore the only method directly comparable to sc2ts.
A recombination event is detected by first
identifying plausible branches carrying multiple mutations.
For each candidate branch,
RIPPLES systematically splits the sequence into 2 or 3 segments, and
finds parsimonious placements for each segment independently.
A recombination event at that node is then flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny
(the ``parsimony improvement'' parameter, $p=3$, by default).
RIVET~\cite{Smith2023tracking} is a tool for quality control and
visualization of these recombination events.

We ran fast-RIPPLES\cite{Smith2023tracking}
on the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
At $p=3$, this resulted in a total of 4111 recombination events.
Most of these subtend $<10$ samples (2616 singletons),
with 4 events subtending more than 100,000 samples.
Analysis of the descending samples for each event shows that
10 of the 44 Pango X lineages present in the dataset
% NOTE: listing them like this for sorting and manipulation purposes
(%
XA, 
XAC, 
XAE, 
XAJ,
XB, 
XBB, 
XBM, 
XC, 
XL, 
XW%
)
are associated unequivocally with a single recombination event.
7 lineages 
(%
XAL,
XAM,
XBD, 
XBR, 
XE,
XS,
XY%
)
are associated with two events. Finally, 
one event subtending 1321 samples suggests that 
XAF,
XAL, 
XE, 
XH, 
XJ, 
XM, 
XY, 
and 
82 BA.2 samples share a recombination event [With what breakpoint and parents?].
We also ran RIPPLES with $p=4$, which reduced the overall number of
events to 1174 (roughly 4-fold) but also approximately halved the number of
events involving X lineages, missing well-characterized events,
such as XBB (Table~\ref{tab:ripples_x_lineages}).

\subsection*{Quality control of recombination events}

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues}
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We developed a quality control (QC) measure
which counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues.
Figure~\ref{fig:recombinant_qc} shows
the net number of supporting loci for the left and right parents
for all 855 recombination events.
The 501 recombinants having fewer than 4 net loci supporting each parent
are highly enriched for samples sequenced using the AmpliSeq V1 protocol
% https://github.com/jeromekelleher/sc2ts-paper/issues/808
(70\% of the recombinants despite making up 2\% of the Viridian dataset), and
71 (14\%) are associated with 6-base deletion in the Delta lineage,
known to be affected by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and
the Jackson et al. recombinants (Document~\ref{sec:jackson_recombs}).
This QC filter also correlates strongly with results of other 
recombination detection methods when run on the putative 
recombinant sequences (see next section).

\subsection*{Validating and evaluating plausibility of recombination events}
[THIS IS VERY MUCH WIP]

We validated the recombination events in the sc2ts ARG by running 
several recombination detection programs on the putative recombinant
sequences. [TODO generate supplemental table summarising results of 
valiation.]

We first ran 3SEQ~\cite{Lam2018improved} [etc etc].

To assess the evidence for these putative novel recombination events we 
ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
90 are classified as 
recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
were classified as non-recombinant).
Rebar classified 203 of the QC-passing sc2ts recombination events, 
as recombinant, with only 10 of those failing QC being classified 
as recombinant.

The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination  (similar to 
idea of ``parsimony improvement'' in RIPPLES.)
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

@@

Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
and the resulting distribution of the number of 
averted mutations (lower histogram)
over all 354 recombination events that pass QC. 
Of these, only 24 
(21 associated with Pango X lineages in Table~\ref{tab:pango_x_lineages},
and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
been thoroughly characterised.  
To assess the evidence for these putative novel recombination events we 
ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
90 are classified as 
recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
were classified as non-recombinant).
Rebar classified 203 of the QC-passing sc2ts recombination events, 
as recombinant, with only 10 of those failing QC being classified 
as recombinant.

[WIP]
Of the 38 recombinants that are strongly supported (at least 10 averted
mutations), 29 are novel. The known recombinants are XA, XC, XS, XBR, 
and 5 from Jackson et al. All but 2 of the 38 are classified as
recombinant by rebar. The majority (20) of these novel strongly supported
recombinants are singletons, with 9 having further descendants in the ARG.
Of the remaining 316 recombinants with less than 10 averted mutations,
only 151 are classified as recombinant by rebar. 
While some are likely to 
be false positives, it is also interesting to note that rebar validation
rate is much lower for recombinants with closely related 
parents (Figure~\ref{fig:recombinant_evidence}B,C) suggesting that 
sc2ts may have a higher sensitivity.

See Figure~\ref{fig:recombinant_evidence_cov_recomb} CovRecomb
See Figure~\ref{fig:recombinant_evidence_rebar} for rebar,
Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p4
and Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p3.

@@

@@

@@

>>>>

\subsection*{Quality control and validation of recombination events}

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues}
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We use a simple quality control (QC) measure
that counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues.
Figure~\ref{fig:recombinant_qc} shows
the net number of supporting loci for the left and right parents
for all 855 recombination events.
The 501 recombinants having fewer than 4 net loci supporting each parent
are highly enriched for samples sequenced using the AmpliSeq V1 protocol
% https://github.com/jeromekelleher/sc2ts-paper/issues/808
(70\% of the recombinants despite making up 2\% of the Viridian dataset), and
71 (14\%) are associated with 6-base deletion in the Delta lineage,
known to be affected by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and
the Jackson et al.\ recombinants (Document~\ref{sec:jackson_recombs}).

We validated the recombination events in the sc2ts ARG by running 
the 3SEQ~\cite{Lam2018improved}, CovRecomb~\cite{Li2024CovRecomb} and rebar 
recombination detection methods on the putative recombinant sequences
(STAR Methods) summarised in Table~\ref{tab:recombinant_validation}.
Overall, validation among the QC-passing recombinants is highest
at with 3SEQ at 88\%, and is lowest for CovRecomb at xx\%. 
CovRecomb has been previously shown to be 
less sensitive than other methods~\cite{Li2024CovRecomb}, 
and is specifically intended 
to detect interlineage recombinants. The 12\% of recombinants 
not passing 3SEQ validation are also the most marginally supported
(see next section; Figure~\ref{fig:recombinant_evidence}).
All of the validation metrics also correlate strongly with the 
QC filter, with very few of the QC-failing recombinants
passing external validation (Table~\ref{tab:recombinant_validation}).

\subsection*{Assessing evidence for recombination events}
The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination  (similar to 
idea of ``parsimony improvement'' in RIPPLES.)
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

@@

Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
and the resulting distribution of the number of 
averted mutations (lower histogram)
over all 354 recombination events that pass QC. 
Also shown is the 3SEQ classification status of these recombinants.
All of the 12\% of recombinants that do not pass 3SEQ validation are 
from the most weakly supported, with 4 or 5 averted mutations.
Figures~\ref{fig:recombinant_evidence_cov_recomb} and
~\ref{fig:recombinant_evidence_rebar} 
same the same information for CovRecomb and rebar validation,
showing that their validation status tends to be function 
of the divergence between the recombinant parents.

@@

\subsection*{Comparison with UShER+RIPPLES}

RIPPLES~\cite{Turakhia2022} is a method for identifying
recombination events using a phylogeny inferred by UShER.
This distinguishes RIPPLES from other recombination detection
methods such as 3SEQ, which classify particular \emph{sequences}
as recombinant or not and can only identify potential onward 
transmission post-hoc. The combination of UShER and RIPPLES
is therefore the only method that is directly comparable 
to sc2ts. It works by first identifying plausible long 
branches in the input tree.
For each candidate branch,
RIPPLES systematically splits the sequence into 2 or 3 segments, and
finds parsimonious placements for each segment independently.
A recombination event at that node is then flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny
(the ``parsimony improvement'' parameter, $p=3$, by default).
RIVET~\cite{Smith2023tracking} is a tool for quality control and
visualization of these recombination events.

To compare the accuracy of sc2ts with UShER+RIPPLES we ran
RIPPLES the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
We used the RIPPLES parameter $p=4$ as is equivalent to the 
sc2ts $k=4$ parameter used to build the ARG, which resulted 
in 11[XX] distinct recombination events. We then ran 3SEQ on
each of the candidate trios, giving an overall validation
rate of XX\% (STAR Methods). Comparing events identified 
in the UShER tree and sc2ts ARG is not straightforward in 
general, and so we focus comparison on the singleton recombinants
returned by both. Although these are likely to be enriched for 
artefacts, both methods face the same challenges and the 
events can be compared like-for like. Of the 11XX recombination
events returned by RIPPLES, 757 are singletons (resp 525/855 for 
sc2ts). Of the RIPPLES singletons, 705 are also singletons 
in the sc2ts ARG. [There's a complicated story to tell here 
about extra singletons in the RIPPLES set which pass
3SEQ validation, but have parents from the future.]
Thus, 216 of these recombinants  pass 3SEQ validation, have 
at least one parent that is on average 63 days \emph{younger}
than the sequence itself. These parents were not available
to sc2ts as potential matches, as it enforces strict 
time ordering based on sample metadata.
Further study is required to determine
if the metadata is incorrect in these cases, or if these 
should be flagged as false positives. However, the 
key point is that RIPPLES is dependent on signals
embedded in the UShER tree, and therefore we cannot 
decouple the finer details of post-hoc recombinant detection
from the tree building algorithm.

We also compared the recombination events output by RIPPLES with
the Pango X lineages. At p=4, we found that only XX of the 
X lineages were present and many well-characterized events 
such as XBB were not present. We therefore reran at p=3,
resulting in XXXX events overall. At p=4, events involving
XX of the 44 Pango X lineages present in the dataset,
with unequivocal origins for 10 lineages, and 
7 lineages other lineages associated with two competing events. 
[More summary]

[Paragraph summing things up]
See Document~\ref{sec:detailed_ripples} for more detailed analysis 
of the UShER+RIPPLES results.

@@

\begin{table}
\caption{Validation of sc2ts recombinants by 3SEQ, CovRecomb and rebar over 
all 855 recombinants (total) and XXX passing and XXX failing QC.
TODO FILL OUT DETAILS.
}
\label{tab:recombinant_validation}
\centering
\begin{tabular}{llll}
\toprule
Method & Overall & QC Pass & QC Fail \\
\midrule
3SEQ & xx & 88\% \\
CovRecomb & \\
rebar & \\
\bottomrule
\end{tabular}
\end{table}

@@

At $p=3$, this resulted in a total of 4111 recombination events.
Most of these subtend $<10$ samples (2616 singletons),
with 4 events subtending more than 100,000 samples.
Analysis of the descending samples for each event shows that
10 of the 44 Pango X lineages present in the dataset
% NOTE: listing them like this for sorting and manipulation purposes
(%
XA, 
XAC, 
XAE, 
XAJ,
XB, 
XBB, 
XBM, 
XC, 
XL, 
XW%
)
are associated unequivocally with a single recombination event.
7 lineages 
(%
XAL,
XAM,
XBD, 
XBR, 
XE,
XS,
XY%
)
are associated with two events. Finally, 
one event subtending 1321 samples suggests that 
XAF,
XAL, 
XE, 
XH, 
XJ, 
XM, 
XY, 
and 
82 BA.2 samples share a recombination event [With what breakpoint and parents?].
We also ran RIPPLES with $p=4$, which reduced the overall number of
events to 1174 (roughly 4-fold) but also approximately halved the number of
events involving X lineages, missing well-characterized events,
such as XBB (Table~\ref{tab:ripples_x_lineages}).

[TODO Discuss  the large effect recombinants inferred by both]
